Canopy leaf area of an evergreen Eucalyptus woodland does not respond to elevated

### 2015-04-26 manuscript.Rmd
More refs and numbers. Added word template (not completely right though)

---
title: Canopy leaf area of an evergreen Eucalyptus woodland does not respond to elevated
  atmospheric [CO$_2$] but tracks water availability
author: "Remko Duursma"
date: "April 24, 2015"
output:
  html_document:
    number_sections: yes
  pdf_document: default
  word_document: default
csl: global-change-biology.csl
bibliography: references.bib
---

@@

# Some pre-calculations
caliblm <- calibrateToDrought(facegap_cloudy_byring)$lmfit

# Some definitions of constants
.gambase <- 18
```

@@

Leaf area, quantified by the leaf area index (*L*), is a crucial driver of forest productivity, water use, and energy balance. Because *L* responds to environmental drivers, it can represent an important feedback to climate change, but its responses to rising [CO$_2$] and water availability of unmanaged forests have been poorly quantified. We studied canopy leaf area dynamics for 27 months in a native evergreen Eucalyptus woodland exposed to Free Air CO$_2$ Enrichment (the EucFACE experiment), in a temperate (hot summer) climate where water limitation is common. We hypothesized that, because of expected higher productivity and increased water-use efficiency, *L* should increase with elevated [CO$_2$]. We estimated *L* from diffuse canopy transmittance and canopy photos, and measured monthly leaf litter production. Contrary to expectation, *L* did not respond to elevated [CO$_2$] in this timeframe. We found that *L* varied between 1.0 and 2.0, but did not show a clear phenological pattern. Instead, increased *L* following leaf flushing occurred a few weeks after heavy rainfall, and *L* decreased steadily during periods of low rainfall. Leaf litter production was correlated to changes in *L*, both during periods of decreasing *L* (when no leaf growth occurred), and during periods of increasing *L* (active shedding of old foliage when new leaf growth occurred). We discuss these results in the context of modelling phenology of evergreen woodlands in response to environmental drivers.

@@

Leaf area index (*L*) is an important ecosystem attribute underlying productivity of vegetation, (Reich, 2012), and a key variable in land surface models (Kala et al., 2013; Walker et al., 2014). *L* is a dynamic canopy property, as it is generally responsive to variation in water and nutrient availability, as demonstrated in experiments (Herbert and Fownes, 1995; Hubbard et al., 2004; White et al., 2010) and observational studies across rainfall gradients (Ellis and Hatton, 2008; Iio et al., 2013) [@Jin2013]. Ecosystem models generally predict that *L* increases in response to elevated atmospheric [CO$_2$] (e*C*$_a$) due to increased net carbon uptake (De Kauwe et al., 2014; Hatton et al., 1992; Kergoat et al., 2002; Zhu et al., 2012) Also expected because increased quantum yield of photosynthesis in e*C*$_a$ would allow leaf retention into deeper shade (Norby et al., 1999), but little empirical evidence is available on the response of *L* to e*C*$_a$ in mature forests (Norby and Zak 2011). Quantifying the magnitude and direction of the e*C*$_a$ effect on *L* is crucial, as changes in *L* associated with climate change drivers such as increases in C$_a$ may provide an important feedback to climate change, and modify direct physiological responses to C$_a$ (Tor-ngern et al., 2015)(Murray et al., 2012).

Direct effects of e*C*$_a$ on vegetation processes include an enhancement of photosynthesis and productivity, a decrease in stomatal conductance and leaf-level water use and an increase in the water-use efficiency (Ainsworth and Long, 2005; Ellsworth et al., 2012). Although numerous studies have been conducted on the response of physiology and plant growth to e*C*$_a$, very few of these inform the response of steady-state *L* to e*C*$_a$, because it requires a sufficiently mature forest to be exposed to free air CO$_2$ enrichment (FACE) (Norby and Zak, 2011). Increased carbon availability through enhanced photosynthesis under e*C*$_a$ does not necessarily lead to increases in *L*. Out of the six forest FACE studies that have been carried out worldwide, only broad-leaved deciduous aspen stands (in monoculture or mixed with birch) and needle-leaved evergreen pine stands showed increased *L* in response to e*C*$_a$ *L* (Norby and Zak, 2011), although positive responses to e*C*$_a$ in terms of productivity were observed at all six sites. Interestingly, Norby and Zak (2011) found a negative correlation between pre-treatment *L* and the responsiveness to e*C*$_a$, leading to the prediction that stands with low *L* will show a larger response to e*C*$_a$. Because diminished returns in increasing LAI in response to e*C*$_a$ when LAI is larger (Norby et al., 2005)

In ecosystems where water availability limits productivity, we could expect e*C*$_a$ to further affect *L* (Donohue et al., 2009), since e*C*$_a$ reduces stomatal conductance, and hence water use per unit of leaf area (Warren et al., 2010) leading to increased soil water availability. These ‘water savings’ can be expected to lead to an increase in *L* in water-limited ecosystems under the idea that leaf area is closely coupled to water-balance. Explain here or just before that we expect this because of the tight relationship between LAI and rainfall across gradients.
In support of this prediction, (Donohue et al., 2013) attributed the observed global increase in *L* in arid and semi-arid areas over the last 30 years to rising C$_a$. Based on the same water-savings principle, we predict that *L* should also increase under e*C*$_a$ in water-limited forests.

@@

We estimated *L* from near-continuous measurements of canopy transmittance in a Eucalyptus-dominated woodland exposed to e*C*$_a$, the EucFACE experiment. The woodland stand at the EucFACE is frequently water-limited, allowing us to quantify the dynamics of *L*, and its components (litter fall and production) in response to water availability. We hypothesized that e*C*$_a$ would result in an increase in *L*, particularly during periods of water limitation, as water savings were expected to lead to longer retention of foliage under elevated CO$_2$. In addition, we expected that the low *L* that is typical of native Eucalyptus woodlands (Whitehead and Beadle, 2004) allowed a larger response to e*C*$_a$ (Norby and Zak, 2011). Our second objective was to quantify the temporal dynamics of *L*, in particular in relation to fluctuating water availability.

@@

Within the EucFACE site, there are six 25-m diameter circular plots (designated hereafter as ‘rings’). Each ring comprises 32 vent pipes in a cylindrical frame extending above the tree canopy to 28 m high. Each vertical pipe is subtended by a control valve and is connected to a circular plenum (Hendrey et al., 1999) through which a computer-controlled system modulates release of diluted CO$_2$ into the vegetation according to a proportional-integral-differential type of algorithm (Lewin et al., 2009). The vegetation within three randomly selected rings was exposed to a CO$_2$ mole fraction (C$_a$) higher than ambient starting in September 2012, whereas the other three rings were identical to the CO$_2$ treatment rings in infrastructure and instrumentation but received ambient C$_a$, and are used as control plots. The target diurnal C$_a$ for the rings exposed to e*C*$_a$ was 150 μmol CO$_2$ mol$^{-1}$ above ambient C$_a$ in full operation mode. However, C$_a$ was increased gradually for six months at a rate of ca. 30 μmol CO$_2$ mol$^{-1}$ month$^{-1}$ until full operation mode was reached (Figure 1a).

@@

## Leaf litter production
In each ring, 8 litter cloth baskets with an area of 0.1979 m$^2$ were installed at 0.5m above the forest floor. Litter was collected approximately monthly, and leaves separated from other components such as twigs, bark, seeds, insect frass, etc. Litter production was converted to units of *L* using an estimate of specific leaf area (SLA) determined from 329 leaves collected from the tree canopies from May 2012 to May 2014 (`r get_sla()` cm$^2$ g$^{-1}$, SE=2.14). We did not find a significant effect of C$_a$ treatment on SLA (Ellsworth, unpublished data).

@@

## Data analysis
We used generalized additive models (gam) for all analyses of *L* and τd timeseries (Wood, 2006), using the mgcv package in R (R Core Team, 2014). For all time-series, we used a base of `r .gambase` when fitting the gams, which resulted in biologically realistic smoothed *L* dynamics. In all fits, we used the FACE ring as a random effect, except when we analysed average *L* across all rings (Fig. 7). We estimated the derivative of the fitted gam, including confidence intervals for the derivative, following Simpson (2014). For comparison between *L* from photos and PAR transmittance, we used simple linear regression. To test for effects of C$_a$ treatment on $L_\tau$ and litter production, we used linear-mixed effects models with FACE ring as the random effect, and C$_a$ treatment and date as a factor.

To assess effects of e*C*$_a$ on $L_\tau$, we accounted for pre-treatment differences in canopy structure, as follows. The average $L_\tau$ over the entire period was well correlated with basal area (BA) measured prior to the onset of the e*C*$_a$ treatment across the FACE rings (Fig. 3). Because the EucFACE woodland is well established, the differences in BA are not due to C$_a$ treatment effects, but instead represent natural variation within the woodland. We then calculated the deviation of actual $L_\tau$ from the fitted relationship with BA shown in Fig. 3, to yield a ‘*L* anomaly’. This way, we effectively used BA as a covariate in the analysis to account for pre-treatment variation.

@@

We found no differences in the *L* anomaly between the C$_a$ treatments (Fig. 4) (p > 0.05), since the 95% confidence intervals of the generalized additive model overlapped. A linear mixed-effects model with C$_a$ and time as fixed effects, and FACE ring as a random effect also showed no significant effect of C$_a$ on the *L* anomaly (P > 0.2). What difference between C$_a$ treatments could we have reasonably detected? The average half-width of the confidence interval on the *L* anomaly (Fig. 4) was 0.047 m$^2$ m$^{-2}$ for a*C*$_a$, and 0.042 m$^2$ m$^{-2}$ for the $eC_a$ treatment. This means that, provided the within-treatment variance stays the same, we could detect a difference of 0.089 m$^2$ m$^{-2}$ (the sum of the half-width CI’s), or about a change of 6% over the mean *L* of the EucFACE (1.39 m$^2$ m$^{-2}$).

@@

We found a significant correlation among FACE rings between litter production (averaged over the entire period) and mean *L*τ during the same period (Fig. 6a), which is expected if leaf lifespan is relatively constant. This allowed us estimate mean leaf lifespan from *L*τ and litter production as the ratio of *L*τ to annual litter production, yielding a mean lifespan of 0.95 yr, with no significant difference (P? F? df?) between C$_a$ treatments (Fig. 6b).

@@

## Response of *L* to elevated [CO$_2$]
We report on 28 months of continuous leaf area index (*L*) measurements in a native evergreen Eucalyptus woodland exposed to +150ppm elevated CO$_2$ (e*C*$_a$) and 3 months of ramp-up in C$_a$. We had predicted that *L* would increase in response to e*C*$_a$ as a result of increased quantum yield for photosynthesis, enhanced CO$_2$ assimilation for the entire canopy, and increased soil water availability. We also expected that this eucalypt woodland, with its inherently low *L* (Whitehead and Beadle, 2004), would be even more responsive to e*C*$_a$ than higher-*L* and more closed-canopy stands (Norby and Zak, 2011). Counter to our expectations but similar to Norby et al. (2003), we found no significant effect of elevated C$_a$ on *L*. It could be argued that the change in *L* in response to e*C*$_a$ was too small to be detected. Natural variation in *L* among plots in a native woodland is indeed a challenge to manipulative studies with low replication, such as FACE experiments. To increase sensitivity and statistical robustness, we used a novel method to account for pre-treatment differences by using stand basal area as a covariate. Using this method, we estimated that we can detect a difference of only a 6% in *L*. Based on this small detectable difference, we conclude that the lack of a significant C$_a$ effect is not due to a lack of power, but a true lack of response to elevated C$_a$.

This lack of a response to elevated C$_a$ counters our expectations and those in previous publications (Norby et al. 2003 McCarthy et al. 2010 etc.), for a number of reasons. First, in a review of forest FACE sites, Norby and Zak (2011) showed that low *L* stands are more responsive to e*C*$_a$. They argued that high *L* canopies would not benefit from further increases in *L*, because it would lead to only very small increases in total absorbed radiation. The mean *L* at the EucFACE was ca. 1.4 m$^2$ m$^{-2}$, which is low compared to a global compendium of forest *L* (Iio et al., 2013), and the lowest *L* of any forest FACE experiment (Norby and Zak, 2011), but in line with a range of values for native Eucalyptus forests and woodlands reported by (Whitehead and Beadle, 2004) and (Iio et al., 2013). Based on this *L* estimate and the relationship across FACE sites reported by Norby and Zak (2011), we could expect about a 30% increase in *L* under e*C*$_a$. It is clearly too early to draw general conclusions across FACE sites in terms of *L* response, moreover because water and nutrient likely play a central role in productivity and *L* dynamics at the EucFACE. Indeed, experimental evidence clearly demonstrates strong limitations of *L* by nutrients and water in Eucalyptus stands (White et al., 2010).

A possible explanation for the lack of a *L* response to e*C*$_a$ is that leaf growth may not be limited by carbon availability. In other words, stimulation of net leaf photosynthesis by e*C*$_a$ (a common response, also observed at the EucFACE, (Gimeno et al. in review) may not lead to increased net leaf growth, due to water or nutrient limitations (Crous et al., 2015). Net leaf growth is the net sum of allocation of photosynthates to leaf growth, and turnover of the leaf pool. Either process may itself be affected by e*C*$_a$ (De Kauwe et al. 2014, refs on turnover). We found no support for differences in leaf lifespan estimates based on *L* and litter production between C$_a$ treatments (Fig. 6), suggesting that during the study period leaf turnover rate was not affected by e*C*$_a$. To further understand the response of biomass allocation to e*C*$_a$ at the EucFACE, all components of the carbon balance must be carefully quantified, a task underway but not yet complete. For example, Drake et al. (in review) reported increased soil respiration under e*C*$_a$, suggesting that at least some of the stimulation of photosynthesis was allocated to either root growth, respiration, or root exudation.

@@

Although the response of *L* to water availability is qualitatively well-known, current-generation phenology models fall short in incorporating this response. In a comparison of six leading ecosystem and global vegetation models,  (Powell et al., 2013) showed that the steady decrease in *L* in an Amazon drought experiment could be matched by only one model. Simulations from De Kauwe et al. (2014) at ambient and elevated C$_a$ also showed little response to inter-annual variation in rainfall (pers. comm. Martin De Kauwe). In most models  that provide prognostic estimates of *L* (i.e., excluding models that take *L* as an input parameter; e.g. (Kala et al., 2013)), temporal changes in *L* are modelled by first calculating total net photosynthetic productivity (carbon availability, gross photosynthesis minus respiration), and allocating some fraction to the leaf pool. Combined with - usually constant - leaf turnover and specific leaf area (ratio of leaf area to mass), this yields predictions of *L* in response to environmental drivers such as e*C*$_a$ (Fig. 2). We hypothesize that this algorithm is not well suited to explain the dynamic nature of *L* in response to water availability observed in evergreen broad-leaved forests, because not all important mechanisms are incorporated in these models. In order to construct a phenology model where phenology is responsive to water availability, it should include responses of both leaf turnover and leaf growth to fluctuating water availability. Water limitation is known to enhance leaf turnover (Pook et al.1997), possibly to maintain sufficient rates of water availability to the remaining leaves (Vilagrosa et al. 2003, Misson et al. 2010). Some existing ecosystem models allow for increased leaf turnover during periods of low water availability (e.g. G’DAY, see Corbeels et al. 2005), but this process has been poorly quantified.

# Conclusions
Leaf area index (*L*) of a natural Eucalyptus woodland was unresponsive to elevated [CO$_2$] after 27 months of treatment and 24 months at +150ppm CO$_2$ elevation. Litter production was closely related to dynamics of *L*, both during dry periods (leading to accelerated litter production), and periods of net leaf growth (when old foliage was replaced by new foliage). We found no clear annual pattern in onset of leaf growth, instead, leaf growth consistently commenced after heavy rainfall following a dry period. Both the dynamics of litter production and the rapid response to soil water availability challenge current-generation prognostic *L* and phenology models. High resolution time series such as the one we present in this study should be used to develop new models of phenology for evergreen woodlands.

>>>>

---
title: Canopy leaf area of an evergreen Eucalyptus woodland does not respond to elevated
  atmospheric [$CO_2$] but tracks water availability
author: "Remko Duursma"
date: "April 24, 2015"
output:
  word_document:
    reference_docx: msstyle.docx
  pdf_document: default
  html_document:
    number_sections: yes
csl: global-change-biology.csl
bibliography: references.bib
---

@@

# Main results
df <- facegap_cloudy_byring

# Some pre-calculations
caliblm <- calibrateToDrought(df)$lmfit

# Number of months of study
r <- range(df$Date)
nummon <- as.numeric(floor(difftime(r[2],r[1],units="weeks")/(52/12)))

# Range mean LAI
dfa <- summaryBy(LAI ~ Date, FUN=mean, na.rm=TRUE, data=df)
rangeLAI <- round(range(dfa$LAI),1)

# Some definitions of constants
.gambase <- 18

```

@@

Leaf area, quantified by the leaf area index (*L*), is a crucial driver of forest productivity, water use, and energy balance. Because *L* responds to environmental drivers, it can represent an important feedback to climate change, but its responses to rising [$CO_2$] and water availability of unmanaged forests have been poorly quantified. We studied canopy leaf area dynamics for `r nummon` months in a native evergreen Eucalyptus woodland exposed to Free Air $CO_2$ Enrichment (the EucFACE experiment), in a temperate (hot summer) climate where water limitation is common. We hypothesized that, because of expected higher productivity and increased water-use efficiency, *L* should increase with elevated [$CO_2$]. We estimated *L* from diffuse canopy transmittance and canopy photos, and measured monthly leaf litter production. Contrary to expectation, *L* did not respond to elevated [$CO_2$] in this timeframe. We found that *L* varied between `r rangeLAI[1]` and `r rangeLAI[2]`, but did not show a clear phenological pattern. Instead, increased *L* following leaf flushing occurred a few weeks after heavy rainfall, and *L* decreased steadily during periods of low rainfall. Leaf litter production was correlated to changes in *L*, both during periods of decreasing *L* (when no leaf growth occurred), and during periods of increasing *L* (active shedding of old foliage when new leaf growth occurred). We discuss these results in the context of modelling phenology of evergreen woodlands in response to environmental drivers.

@@

Leaf area index (*L*) is an important ecosystem attribute underlying productivity of vegetation, [@Reich2012], and a key variable in land surface models [@Kala2013; @Walker2014]. *L* is a dynamic canopy property, as it is generally responsive to variation in water and nutrient availability, as demonstrated in experiments [@Herbert1995; @Hubbard2004; @White2010] and observational studies across rainfall gradients [@Ellis2008; @Iio2013; @Jin2013]. Ecosystem models generally predict that *L* increases in response to elevated atmospheric [$CO_2$] ($eC_a$) due to increased net carbon uptake [@DeKauwe2014; @Hatton1992; @Kergoat2002; @Zhu2012]. Also expected because increased quantum yield of photosynthesis in $eC_a$ would allow leaf retention into deeper shade [@Norby1999], but little empirical evidence is available on the response of *L* to $eC_a$ in mature forests [@Norby2011]. Quantifying the magnitude and direction of the $eC_a$ effect on *L* is crucial, as changes in *L* associated with climate change drivers such as increases in $C_a$ may provide an important feedback to climate change, and modify direct physiological responses to $C_a$ [@Tor-ngern2015; @Murray2012].

Direct effects of $eC_a$ on vegetation processes include an enhancement of photosynthesis and productivity, a decrease in stomatal conductance and leaf-level water use and an increase in the water-use efficiency (Ainsworth and Long, 2005; Ellsworth et al., 2012). Although numerous studies have been conducted on the response of physiology and plant growth to $eC_a$, very few of these inform the response of steady-state *L* to $eC_a$, because it requires a sufficiently mature forest to be exposed to free air $CO_2$ enrichment (FACE) (Norby and Zak, 2011). Increased carbon availability through enhanced photosynthesis under $eC_a$ does not necessarily lead to increases in *L*. Out of the six forest FACE studies that have been carried out worldwide, only broad-leaved deciduous aspen stands (in monoculture or mixed with birch) and needle-leaved evergreen pine stands showed increased *L* in response to $eC_a$ *L* (Norby and Zak, 2011), although positive responses to $eC_a$ in terms of productivity were observed at all six sites. Interestingly, Norby and Zak (2011) found a negative correlation between pre-treatment *L* and the responsiveness to $eC_a$, leading to the prediction that stands with low *L* will show a larger response to $eC_a$. Because diminished returns in increasing LAI in response to $eC_a$ when LAI is larger (Norby et al., 2005)

In ecosystems where water availability limits productivity, we could expect $eC_a$ to further affect *L* (Donohue et al., 2009), since $eC_a$ reduces stomatal conductance, and hence water use per unit of leaf area (Warren et al., 2010) leading to increased soil water availability. These ‘water savings’ can be expected to lead to an increase in *L* in water-limited ecosystems under the idea that leaf area is closely coupled to water-balance. Explain here or just before that we expect this because of the tight relationship between LAI and rainfall across gradients.
In support of this prediction, (Donohue et al., 2013) attributed the observed global increase in *L* in arid and semi-arid areas over the last 30 years to rising $C_a$. Based on the same water-savings principle, we predict that *L* should also increase under $eC_a$ in water-limited forests.

@@

We estimated *L* from near-continuous measurements of canopy transmittance in a Eucalyptus-dominated woodland exposed to $eC_a$, the EucFACE experiment. The woodland stand at the EucFACE is frequently water-limited, allowing us to quantify the dynamics of *L*, and its components (litter fall and production) in response to water availability. We hypothesized that $eC_a$ would result in an increase in *L*, particularly during periods of water limitation, as water savings were expected to lead to longer retention of foliage under elevated $CO_2$. In addition, we expected that the low *L* that is typical of native Eucalyptus woodlands (Whitehead and Beadle, 2004) allowed a larger response to $eC_a$ (Norby and Zak, 2011). Our second objective was to quantify the temporal dynamics of *L*, in particular in relation to fluctuating water availability.

@@

Within the EucFACE site, there are six 25-m diameter circular plots (designated hereafter as ‘rings’). Each ring comprises 32 vent pipes in a cylindrical frame extending above the tree canopy to 28 m high. Each vertical pipe is subtended by a control valve and is connected to a circular plenum (Hendrey et al., 1999) through which a computer-controlled system modulates release of diluted $CO_2$ into the vegetation according to a proportional-integral-differential type of algorithm (Lewin et al., 2009). The vegetation within three randomly selected rings was exposed to a $CO_2$ mole fraction ($C_a$) higher than ambient starting in September 2012, whereas the other three rings were identical to the $CO_2$ treatment rings in infrastructure and instrumentation but received ambient $C_a$, and are used as control plots. The target diurnal $C_a$ for the rings exposed to $eC_a$ was 150 μmol $CO_2$ mol$^{-1}$ above ambient $C_a$ in full operation mode. However, $C_a$ was increased gradually for six months at a rate of ca. 30 μmol $CO_2$ mol$^{-1}$ month$^{-1}$ until full operation mode was reached (Figure 1a).

@@

## Leaf litter production
In each ring, 8 litter cloth baskets with an area of 0.1979 m$^2$ were installed at 0.5m above the forest floor. Litter was collected approximately monthly, and leaves separated from other components such as twigs, bark, seeds, insect frass, etc. Litter production was converted to units of *L* using an estimate of specific leaf area (SLA) determined from 329 leaves collected from the tree canopies from May 2012 to May 2014 (`r get_sla()` cm$^2$ g$^{-1}$, SE=2.14). We did not find a significant effect of $C_a$ treatment on SLA (Ellsworth, unpublished data).

@@

## Data analysis
We used generalized additive models (gam) for all analyses of *L* and τd timeseries (Wood, 2006), using the mgcv package in R (R Core Team, 2014). For all time-series, we used a base of `r .gambase` when fitting the gams, which resulted in biologically realistic smoothed *L* dynamics. In all fits, we used the FACE ring as a random effect, except when we analysed average *L* across all rings (Fig. 7). We estimated the derivative of the fitted gam, including confidence intervals for the derivative, following Simpson (2014). For comparison between *L* from photos and PAR transmittance, we used simple linear regression. To test for effects of $C_a$ treatment on $L_\tau$ and litter production, we used linear-mixed effects models with FACE ring as the random effect, and $C_a$ treatment and date as a factor.

To assess effects of $eC_a$ on $L_\tau$, we accounted for pre-treatment differences in canopy structure, as follows. The average $L_\tau$ over the entire period was well correlated with basal area (BA) measured prior to the onset of the $eC_a$ treatment across the FACE rings (Fig. 3). Because the EucFACE woodland is well established, the differences in BA are not due to $C_a$ treatment effects, but instead represent natural variation within the woodland. We then calculated the deviation of actual $L_\tau$ from the fitted relationship with BA shown in Fig. 3, to yield a ‘*L* anomaly’. This way, we effectively used BA as a covariate in the analysis to account for pre-treatment variation.

@@

We found no differences in the *L* anomaly between the $C_a$ treatments (Fig. 4) (p > 0.05), since the 95% confidence intervals of the generalized additive model overlapped. A linear mixed-effects model with $C_a$ and time as fixed effects, and FACE ring as a random effect also showed no significant effect of $C_a$ on the *L* anomaly (P > 0.2). What difference between $C_a$ treatments could we have reasonably detected? The average half-width of the confidence interval on the *L* anomaly (Fig. 4) was 0.047 m$^2$ m$^{-2}$ for $aC_a$, and 0.042 m$^2$ m$^{-2}$ for the $eC_a$ treatment. This means that, provided the within-treatment variance stays the same, we could detect a difference of 0.089 m$^2$ m$^{-2}$ (the sum of the half-width CI’s), or about a change of 6% over the mean *L* of the EucFACE (1.39 m$^2$ m$^{-2}$).

@@

We found a significant correlation among FACE rings between litter production (averaged over the entire period) and mean *L*τ during the same period (Fig. 6a), which is expected if leaf lifespan is relatively constant. This allowed us estimate mean leaf lifespan from *L*τ and litter production as the ratio of *L*τ to annual litter production, yielding a mean lifespan of 0.95 yr, with no significant difference (P? F? df?) between $C_a$ treatments (Fig. 6b).

@@

## Response of *L* to elevated [$CO_2$]
We report on 28 months of continuous leaf area index (*L*) measurements in a native evergreen Eucalyptus woodland exposed to +150ppm elevated $CO_2$ ($eC_a$) and 3 months of ramp-up in $C_a$. We had predicted that *L* would increase in response to $eC_a$ as a result of increased quantum yield for photosynthesis, enhanced $CO_2$ assimilation for the entire canopy, and increased soil water availability. We also expected that this eucalypt woodland, with its inherently low *L* (Whitehead and Beadle, 2004), would be even more responsive to $eC_a$ than higher-*L* and more closed-canopy stands (Norby and Zak, 2011). Counter to our expectations but similar to Norby et al. (2003), we found no significant effect of elevated $C_a$ on *L*. It could be argued that the change in *L* in response to $eC_a$ was too small to be detected. Natural variation in *L* among plots in a native woodland is indeed a challenge to manipulative studies with low replication, such as FACE experiments. To increase sensitivity and statistical robustness, we used a novel method to account for pre-treatment differences by using stand basal area as a covariate. Using this method, we estimated that we can detect a difference of only a 6% in *L*. Based on this small detectable difference, we conclude that the lack of a significant $C_a$ effect is not due to a lack of power, but a true lack of response to elevated $C_a$.

This lack of a response to elevated $C_a$ counters our expectations and those in previous publications (Norby et al. 2003 McCarthy et al. 2010 etc.), for a number of reasons. First, in a review of forest FACE sites, Norby and Zak (2011) showed that low *L* stands are more responsive to $eC_a$. They argued that high *L* canopies would not benefit from further increases in *L*, because it would lead to only very small increases in total absorbed radiation. The mean *L* at the EucFACE was ca. 1.4 m$^2$ m$^{-2}$, which is low compared to a global compendium of forest *L* (Iio et al., 2013), and the lowest *L* of any forest FACE experiment (Norby and Zak, 2011), but in line with a range of values for native Eucalyptus forests and woodlands reported by (Whitehead and Beadle, 2004) and (Iio et al., 2013). Based on this *L* estimate and the relationship across FACE sites reported by Norby and Zak (2011), we could expect about a 30% increase in *L* under $eC_a$. It is clearly too early to draw general conclusions across FACE sites in terms of *L* response, moreover because water and nutrient likely play a central role in productivity and *L* dynamics at the EucFACE. Indeed, experimental evidence clearly demonstrates strong limitations of *L* by nutrients and water in Eucalyptus stands (White et al., 2010).

A possible explanation for the lack of a *L* response to $eC_a$ is that leaf growth may not be limited by carbon availability. In other words, stimulation of net leaf photosynthesis by $eC_a$ (a common response, also observed at the EucFACE, (Gimeno et al. in review) may not lead to increased net leaf growth, due to water or nutrient limitations (Crous et al., 2015). Net leaf growth is the net sum of allocation of photosynthates to leaf growth, and turnover of the leaf pool. Either process may itself be affected by $eC_a$ (De Kauwe et al. 2014, refs on turnover). We found no support for differences in leaf lifespan estimates based on *L* and litter production between $C_a$ treatments (Fig. 6), suggesting that during the study period leaf turnover rate was not affected by $eC_a$. To further understand the response of biomass allocation to $eC_a$ at the EucFACE, all components of the carbon balance must be carefully quantified, a task underway but not yet complete. For example, Drake et al. (in review) reported increased soil respiration under $eC_a$, suggesting that at least some of the stimulation of photosynthesis was allocated to either root growth, respiration, or root exudation.

@@

Although the response of *L* to water availability is qualitatively well-known, current-generation phenology models fall short in incorporating this response. In a comparison of six leading ecosystem and global vegetation models,  (Powell et al., 2013) showed that the steady decrease in *L* in an Amazon drought experiment could be matched by only one model. Simulations from De Kauwe et al. (2014) at ambient and elevated $C_a$ also showed little response to inter-annual variation in rainfall (pers. comm. Martin De Kauwe). In most models  that provide prognostic estimates of *L* (i.e., excluding models that take *L* as an input parameter; e.g. (Kala et al., 2013)), temporal changes in *L* are modelled by first calculating total net photosynthetic productivity (carbon availability, gross photosynthesis minus respiration), and allocating some fraction to the leaf pool. Combined with - usually constant - leaf turnover and specific leaf area (ratio of leaf area to mass), this yields predictions of *L* in response to environmental drivers such as $eC_a$ (Fig. 2). We hypothesize that this algorithm is not well suited to explain the dynamic nature of *L* in response to water availability observed in evergreen broad-leaved forests, because not all important mechanisms are incorporated in these models. In order to construct a phenology model where phenology is responsive to water availability, it should include responses of both leaf turnover and leaf growth to fluctuating water availability. Water limitation is known to enhance leaf turnover (Pook et al.1997), possibly to maintain sufficient rates of water availability to the remaining leaves (Vilagrosa et al. 2003, Misson et al. 2010). Some existing ecosystem models allow for increased leaf turnover during periods of low water availability (e.g. G’DAY, see Corbeels et al. 2005), but this process has been poorly quantified.

# Conclusions
Leaf area index (*L*) of a natural Eucalyptus woodland was unresponsive to elevated [$CO_2$] after 27 months of treatment and 24 months at +150ppm $CO_2$ elevation. Litter production was closely related to dynamics of *L*, both during dry periods (leading to accelerated litter production), and periods of net leaf growth (when old foliage was replaced by new foliage). We found no clear annual pattern in onset of leaf growth, instead, leaf growth consistently commenced after heavy rainfall following a dry period. Both the dynamics of litter production and the rapid response to soil water availability challenge current-generation prognostic *L* and phenology models. High resolution time series such as the one we present in this study should be used to develop new models of phenology for evergreen woodlands.
